Practical Machine Learning Course Project
========================================================

Introduction
-------------------------------

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).



Data
====


The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

you need the following libraries

```{r, eval=FALSE}
library(caret)
library(parallel)
library(doParallel)
#load raw data
training <- read.csv("pml-training.csv", header = TRUE)
test  <- read.csv('pml-testing.csv')
```



Cleaning data sets
-------------------------------

I need do the following activities after the training data
- Removing zero variance predictors
- Removing columns about 90% of not a number
- remove the next  columns for classification (x, user name, raw time stamp 1 and 2, new window and num window).
- Convert class into factor

```{r, eval=FALSE}
#remove columns 
nasPerColumn<- apply(training,2,function(x) {sum(is.na(x))});
training <- training[,which(nasPerColumn <  nrow(training)*0.9)];  
  
#remove predictors
nearZeroColumns <- nearZeroVar(training, saveMetrics = TRUE)
training <- training[, nearZeroColumns$nzv==FALSE]
  
#remove not relevant columns for classification (x, user_name, raw time stamp 1  and 2, "new_window" and "num_window")
training<-training[,7:ncol(training)]
  
#class into factor
training$classe <- factor(training$classe)
```


Splitting data
-------------------------------
Split the data: 60% for training, 40% for testing

```{r, eval=FALSE}
trainIndex <- createDataPartition(y = training$classe, p=0.6,list=FALSE);
trainingPartition <- training[trainIndex,];
testingPartition <- training[-trainIndex,];
```

Create machine learning models
-------------------------------
Three models are generated:  random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda") model. Parallel computing methods are employed to improve efficiency. The Principal Component Analysis (PCA) can also be added as a preprocess option of the train function, but at the expense of losing accuracy.

```{r, eval=FALSE}
#random
set.seed(3433)
registerDoParallel(makeCluster(detectCores()))
#three models:
model_rf <- train(classe ~ .,  method="rf", data=trainingPartition)    
model_gbm <-train(classe ~ ., method = 'gbm', data = trainingPartition)
model_lda <-train(classe ~ ., method = 'lda', data = trainingPartition) 
```

Analysis accuracy
-------------------------------
 the accuracy returned for random forest, boosted trees and linear discriminant analysis are 99%, 96% and 69\%. Therefore, the random forest model is the most promising model to be optimized with cross validation.  This was the expected output since random forest was the most complex technique evaluated because it combines a number of classifiers. 

```{r, eval=FALSE}
print("Random forest accuracy ")
rf_accuracy<- predict(model_rf, testingPartition)
print(confusionMatrix(rf_accuracy, testingPartition$classe))
print("Boosted trees accuracy ")
gbm_accuracy<- predict(model_gbm , testingPartition)
print(confusionMatrix(gbm_accuracy, testingPartition$classe))
print("Linear discriminant analysis")
lda_accuracy<- predict(model_lda , testingPartition)
print(confusionMatrix(lda_accuracy, testingPartition$classe))
```

Tuning with cross validation
-------------------------------
To improve the model obtained and more specifically to avoid over-fitting, the cross validation technique is employed with 10 folds.

```{r, eval=FALSE}
#random seed
set.seed(3433)
#parallel computing for multi-core
registerDoParallel(makeCluster(detectCores()))  
controlf <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
model_rf_CV <- train(classe ~ ., method="rf",  data=trainingPartition, trControl = controlf)
```

model cross validation 

```{r, eval=FALSE}
print("Random forest accuracy after CV")
rf_CV_accuracy<<- predict(model_rf_CV , testingPartition)
print(confusionMatrix(rf_CV_accuracy, testingPartition$classe))
```

Accuracy = 0.9921, which is only slightly better than the baseline accuracy (0.9916) which was already very high 


Variables importance
-------------------------------
the final model, the importance of the variables in this model can be studied. 


```{r, eval=FALSE}
print("Variables importance in model")
vi = varImp(model_rf_CV$finalModel)
vi$var<-rownames(vi)
vi = as.data.frame(vi[with(vi, order(vi$Overall, decreasing=TRUE)), ])
rownames(vi) <- NULL
print(vi)
```



Predicting 20 test cases
-------------------------------
Finally, the random forest model tuned with cross validation (model_rf_CV) is used to predict  20 test cases available in the test data loaded at the beginning of the project.

```{r, eval=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}


#Prediction Assignment Submission
predictionassignmet<- function(){
  prediction <- predict(model_rf_CV, test)
  print(prediction)
  answers <- as.vector(prediction)
  pml_write_files(answers)
}

```

Conclusions
-------------------------------
The above code concludes that *roll_belt* is the most important predictor for the model obtained with random forest and tuned by cross validation.
